---
editor: 
  markdown: 
    wrap: 72
---

# Loading and Analyzing Data {#r-load-analyze-visualize}

This tutorial, based on my 2024 presentation for the National Institute
for Computer Assisted Reporting (NICAR), will show you the essential
workflow for any data analysis project in R. You will learn to import
data, explore it, use basic commands to sort and filter and build
summary tables. You will build a basic data visualization from your work
-- all in R code.

You will be executing commands in this document that are contained in
"chunks," which are separate from the text and contain live R code.
**Click the green arrow at the right on line 6** and run the
help.start() command

```{r}
help.start()
```

<br> <br> In the bottom right window of R Studio, you will see a Help
window that displays basic help commands for the program. 

### Install software to grab data

```         
Tidyverse: Eight separate software packages to perform
data import, tidying, manipulation, visualisation, and            programming

Rio: Easy importing features 
Janitor: Data cleaning
```

You should have installed tidyverse already. If not, then delete the
hashtag in front of install.packages("tidyverse") and run the code chunk
at line 22.

```{r}
options(repos = "https://cloud.r-project.org") 
install.packages("tidyverse")
install.packages("rio")
install.packages("janitor")
```

Remember, package installation usually is a one-time thing on your hard
drive. But when you need to load the software libraries each time you
start a script. Libraries are bits of software you will have to load
each time into R to make things run.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(rio)
library(janitor)
```

Check to see what's installed by clicking on "Packages" tab in File
Manager, lower right pane <br>

### Data

We will work with a dataset of MediaSalaries that I cleaned and modified
slightly for this tutorial. Some of the detail has been removed so we
can make calculations. This MediaSalaries sheet was a crowdsourced
project involving reporters worldwide to share information about
salaries and benefits. Open this [file in Google
Sheets](https://docs.google.com/spreadsheets/d/1jkbQFwIdaWv8K00Ad6Wq7ZxFTUPFQA-g/edit#gid=1655992481)

```         
1) Select Salaries tab

IRE Old School: Four Corners Test!
13 Columns
1658 Rows

Numberic data in Salary, Years Experience
Mixed string data in Gender Identity / Ethnicity, Job duties
```

<br> <br>

### Import Data

```         
We'll now load this data into R. You can load spreadsheets from the Internet as follows:
```

```{r}
MediaBucks <- rio::import("https://docs.google.com/spreadsheets/d/1jkbQFwIdaWv8K00Ad6Wq7ZxFTUPFQA-g/edit#gid=1655992481", which = "RealMediaSalaries2")
```

<br> What happened? Look at the table

```{r}
View(MediaBucks)
```

What happened?

R grabbed the spreadsheet from the folder\
We told R to grab the first sheet, RealMediaSalaries2\
R created a dataframe called MediaBucks\
basics of R: \<- is known as an "assignment operator."\
It means: "Make the object named to the left equal to the output of the
code to the right."

<br> <br>

### Explore Data

**Click the green arrow code chunk to get the answers below.**

How many rows?

```{r}
nrow(MediaBucks)
```

How many columns?

```{r}
ncol(MediaBucks)
```

Dimensions: Gives number rows, then columns

```{r}
dim(MediaBucks)
```

Names of your columns

```{r}
colnames(MediaBucks)
```

OR

```{r}
names(MediaBucks)
```

Check data types

```{r}
str(MediaBucks)
```

Let's look at the first six rows

```{r}
head(MediaBucks)
```

Here is a quick way to view the range of your data

```{r}
summary(MediaBucks$Salary)
```

Size and scope

```{r}
sum(MediaBucks$Salary, na.rm=TRUE)
```

\$106 million! for 1,658 journalists

Context: NYT earnings in 2020 = \$100 m Facebook profit for one day:
\$114 million (Q42021=\$10.3B)

average

```{r}
mean(MediaBucks$Salary, na.rm=TRUE)
```

Distribution

```{r}
quantile(MediaBucks$Salary, c(0.1, 0.2, 0.3, 0.4,0.5, 0.6, 0.7, 0.8, 0.9), na.rm=TRUE)
```

```{r}
quantile(MediaBucks$Salary, c(0.25, 0.50, 0.75, 0.9, 0.99), na.rm=TRUE)
```

<br>

### Navigation Tips

```         
Shortcut Commands
Tab - Autocomplete
In Console Window (lower left) 
--Control (or Command) + UP arrow - last lines run
Control (or Command) + Enter - Runs current or selected lines of code in the top left box of RStudio
Shift + Control (or Command) +P - Reruns previous region code
```

<br> <br>

### Dplyr

```         
dplyr has many tools for data analysis   
select Choose which columns to include  
filter Filter the data 
arrange Sort the data, by size for continuous variables, by date, or alphabetically 
group_by Group the data by a categorical variable 
```

<br>

Build a simple summary table by Gender

```{r}
MediaBucks %>% 
  select(Gender, Salary) %>% 
  group_by(Gender) %>% 
  summarize(Total = sum(Salary, na.rm=TRUE))
```

What is the sample size?

```{r}
MediaBucks %>% 
  count(Gender) %>% 
  arrange(desc(n))
```

Better idea: Check Averages!

Build a simple summary table by Gender

```{r}
MediaBucks %>% 
  select(Gender, Salary) %>% 
  group_by(Gender) %>% 
  summarize(Avg_Salary = mean(Salary, na.rm=TRUE))
```

Quick filter out hourly workers

```{r}
MediaSalary <- MediaBucks %>% 
  filter(Salary >= 1000)
```

Just give me a list of the top 10 salaries and companies: use slice_max.
slice_max and slice_min are features in the Dplyr library (part of
Tidyverse) that produce quick summary tables. See what else you can do
with [the slice
commands.](https://dplyr.tidyverse.org/reference/slice.html)

```{r}
MediaBucks %>% 
  select(COMPANY, Salary) %>% 
  slice_max(Salary, n = 10)

```

Questions:

```         
1: View the range of your data    
2: Number of rows  
3: Number of rows cut with filter  
```

<br>

### Find Your News Organization

Filter

```{r}
WSJ <- subset(MediaBucks, COMPANY=="WallStreetJournal")  
```

```{r}
summary(WSJ$Salary)
```

Using Wildcards

```{r}
Journal <- subset(MediaBucks, grepl("?Journal", COMPANY))
```

```{r}
Bloom <- subset(MediaBucks, grepl("?Bloomberg", COMPANY))
```

<br> \### More Tables

Build a table with several companies of your choice

```{r}
BigBoys <- filter(MediaSalary, COMPANY %in% c("NewYorkTimes", "WallStreetJournal", "Bloomberg"))    
```

Table with just reporter salaries

```{r}
Reporters <- subset(MediaBucks, grepl("?reporter", TITLE))
summary(Reporters$Salary)
```

Questions:

```         
1: Who is making $230,504 as a reporter???   
2: Make a table for editors, figure out medians.   
3: Find highest paid editor. Resent them.   
4: Make a table for any position involving data  
```

Table with Black reporters at Wall Street Journal

```{r}
WSJ_Black <- MediaBucks %>% filter(Race =="black", COMPANY == "WallStreetJournal")
```

Build a simple summary table by Race

```{r}
Race <- MediaBucks %>% 
  select(Race, Salary) %>% 
  group_by(Race) %>% 
  summarize(Avg_Salary = mean(Salary, na.rm=TRUE)) %>% 
  arrange(desc(Avg_Salary))
Race
```

Wait! What are the totals by race?

```{r}
MediaBucks %>% 
  count(Race) %>% 
  arrange(desc(n))
```

Advanced: Build a summary table and count by race

```{r}
MediaBucks %>% 
  select(Race, Salary) %>% 
 group_by(Race) %>% 
  summarize(Total=n(),
            Avg = mean(Salary, na.rm=TRUE)) %>% 
  arrange(desc(Total))

#details: https://stackoverflow.com/questions/36183601/average-and-count-with-aggregation-in-r-with-dplyr
```

### Visualize

Let's make a simple chart of our salaries by race.

```{r}
Race %>% 
ggplot(aes(x = Race, y = Avg_Salary, fill = Avg_Salary)) +
  geom_col(position = "dodge") + 
  theme(legend.position = "none") +
  labs(title = "Sample chart of race and salary using MediaSalaries", 
       caption = "1658 records, sample data. Graphic by Rob Wells, 8/11/2022",
       y="Average Salary",
       x="Race / Ethnicity")


```

This is a basic chart using ggplot. To break down the code: - These
lines: Race %\>% ggplot(aes(x = Race, y = Avg_Salary, fill =
Avg_Salary)) + -- Uses the Race table, calls the ggplot program, assigns
the x axis to Race, y axis to Avg_Salary and fills the color according
to Avg_Salary - These lines: geom_col(position = "dodge") +
theme(legend.position = "none") + -- creates a chart of columns, and
removes a legend box. - These lines: labs(title = Assign the headline
and captions.

### What You Have Learned So Far

```         
How to navigate in R studio   
How to install libraries and packages    
How to import a .xlsx file into R   
How to obtain summary statistics (summary)   
How to build basic tables from a dataset   
How to conduct filter queries from a dataset   
```

### Questions

```         
1: Build a table for NewYorkTimes employees, and determine median salary of NewYorkTimes employees   
```

```{r}
#your answer here
```

```         
  2: Identify title, gender and race of the highest paid position at NYT  
```

```{r}
#your answer here
```

```         
3: Search for Bloomberg,  check median salary, compare to NYT results above.
```

```{r}
#your answer here

```

### Tutorials

MaryJo Webster tutorials
<https://sites.google.com/view/mj-basic-data-academy/intro-to-r?authuser=0>

Merrill College eBook: Data Journalism with R and the Tidyverse
<https://wellsdata.github.io/data_journalism_class/_book/>

Excellent book by Sharon Machlis
<https://www.routledge.com/Practical-R-for-Mass-Communication-and-Journalism/Machlis/p/book/9781138726918>

First five chapters are free on her website. My recommendation: buy the
book. <https://www.machlis.com/R4Journalists/>

All Cheat Sheets <https://www.rstudio.com/resources/cheatsheets/>

Andrew Ba Tran first Data Analysis Steps Using R
<https://docs.google.com/presentation/d/1O0eFLypJLP-PAC63Ghq2QURAnhFo6Dxc7nGt4y_l90s/edit#slide=id.p>

Base R Cheat Sheet
<https://www.povertyactionlab.org/sites/default/files/r-cheat-sheet.pdf>
