#Tutorial, April 16, 2024


```{r}
library(tidyverse)
library(janitor)

###
# tidycensus
###
library(tidycensus)

```


```{r}
census_api_key("xyz", install=TRUE, overwrite=TRUE)
```

```{r}

###
# Load ACS variables
###

acs_vars <- load_variables("acs5", year=2020)

###
# Get median household income for Maryland 2020
###

md <- get_acs(geography = "county",
              variables = c(median_income = "B19013_001"),
              state = "MD",
              year = 2020) 


```

### Develop and answer questions

**Task** Develop two different *interesting* questions that you can answer with the data. Run them by your professor before getting started on answering, to ensure they're sufficiently interesting and possible with the data.

```{r}
#Which state had the largest increase in Asian population?
#B02001_005
#Which state had the largest decrease in foreign born population?
#B05002_013
```

**Task** Write code to answer the questions. Write up a sentence with each of your findings. Produce at least one visualization to support one of your findings.

```{r}
#Answering the first question
#Create data sets 
asian_2020 <- get_acs(geography = "state",
              variables = c(asian_divide = "B02001_005"),
              year = 2020) %>% 
  rename(estimate_2020 = estimate)

asian_2010 <- get_acs(geography = "state",
              variables = c(asian_divide = "B02001_005"),
              year = 2010) %>% 
  rename(estimate_2010 = estimate)
#Join the two data frames together
#rsw comment: your version led to a sloppy join:
#Join the two data frames together
total_asian <- asian_2010 %>%
  inner_join(asian_2020, by="GEOID")

#rsw comment: you don't want to do this:
#(estimate.y-estimate.x)/(estimate.x)
#it looks awful and leads to confusion
total_asian <-total_asian %>% 
  mutate(change_asian = (estimate.y-estimate.x)/(estimate.x)*100) %>% 
  arrange(desc(change_asian))

#this version tightens up the df and makes it much more professional

total_asian2 <- asian_2010 %>%
  inner_join(asian_2020, by=c("GEOID", "NAME")) %>% 
  select(GEOID, NAME, estimate_2010, estimate_2020)

#then you get a cleaner calculation

total_asian2 <-total_asian2 %>% 
  mutate(change_asian = (estimate_2020-estimate_2010)/(estimate_2010)*100) %>% 
  arrange(desc(change_asian))

#North Dakota had the largest increase in Asian population between 2010 to 2020, going from a population of 6132 to 11979 which is a 95.4% increase. 


```


```{r}
# Correct answers. I wound up doing it a more efficient way by using the ACS total population estimate. The numbers are more up to date and it throws off the calculations a bit, but it has the same basic answer:

male <- get_acs(geography = "state",
                variables = c(male_population = "B01001_002", total_pop = "B01001_001"),
                year = 2020) %>% 
  rename(state = NAME) %>% 
  select(state, variable, estimate) %>% 
  pivot_wider(names_from = variable, values_from = estimate) %>% 
  mutate(pct_male = male_population/total_pop *100) %>% 
  arrange(desc(pct_male))



```


```{r}
#Answering the second question
foreign_born_2010 <- get_acs(geography = "state",
              variables = c(foreign_born = "B05002_013"),
              year = 2010) 
foreign_born_2020 <- get_acs(geography = "state",
              variables = c(foreign_born = "B05002_013"),
              year = 2020) 
total_foreign_born <- foreign_born_2010 %>%
  inner_join(foreign_born_2020, by="GEOID")

total_foreign_born <-total_foreign_born %>% 
  mutate(change_foreign = (estimate.y-estimate.x)/(estimate.x)*100) %>% 
  arrange(desc(change_foreign))

#Puerto Rico had the largest decrease in foreign born population from 2010 to 2020, going from 105593 to 88386, a -16.3% decrease. 


```

```{r}

white_youth_pop <- get_acs(
  geography = "state",
  variables = c(
    population = "B01001A_003", "B01001A_004", "B01001A_005", "B01001A_006",
    "B01001A_018", "B01001A_019", "B01001A_020", "B01001A_021"
  ),
  year = 2020
)

black_youth_pop <- get_acs(
  geography = "state",
  variables = c(
    population = "B01001B_003", "B01001B_004", "B01001B_005", "B01001B_006",
    "B01001B_018", "B01001B_019", "B01001B_020", "B01001B_021"
  ),
  year = 2020
)

total_white_youth_pop <- sum(white_youth_pop$estimate)
total_black_youth_pop <- sum(black_youth_pop$estimate)



total_pop <- get_acs(geography = "state",
              variables = c(total_population = "B01001_001", black_male_population = "B01001B_001"),
              year = 2020) %>%
  arrange(desc(estimate))
```

```{r}
black_pop <- get_acs(geography = "state",
              variables = c(male_population = "B01001B_001"),
              year = 2020) %>%
  arrange(desc(estimate))
```

```{r}
joined_black_total_pop <- total_pop %>%
  inner_join(black_pop, by="NAME")

```

