---
title: "Example Quarto Markdown HTML"
author: "Sean Mussenden"
date: "`r Sys.Date()`"
format:
  html:
    theme: cosmo
    layout: full
    self-contained: true
    toc: true
    toc-depth: 1
    code-fold: false
execute: 
  echo: true
  warning: false
  message: false
---

## Summary

This document provides an exploration of possibilities with Quarto markdown. Pay attention to the YAML (at the top), the markdown (to style in the text areas), and the special properies (echo, eval) inside of the code blocks.

**Task** You should render it **right now**, and take a look at the output html file. 

**Task**: Next time through, **make changes** described in each section and look at how it changes things.


## Yaml explanation

At the top of the document, you'll see YAML settings, which control the output of the html document.  There are some new ones here, notably code-fold, which collapses all of the code, but allows users to click a toggle to expose it and toc, which loads a table of contents, based on the markdown headings.  

#### Yaml changes next time through

**Task** on your second render, make these changes.
* Change the author name to your name.
* Pick a theme from [this page](https://quarto.org/docs/output-formats/html-themes.html) and replace `cosmo` above with a style one you like. 
* Change toc-depth from 1 to 2. 
* Change code-fold to true

**Task**: What did you expect to happen? Describe here what happened when you made these changes. 

## Codeblock explanation

We've used codeblock QMD properties (eval, echo, warning) to control what's output. Notice the first time through what's included and what's run and what's not.

Here's an example of code that shouldn't produce any output, but we should be able to see the gray box showing the code. 

```{r}
#| echo: true
#| eval: false

print("Set this way, the code should display, but the output should not show up on render")

```

Here's an example of code that we shouldn't be able to see the gray box showing the code, but we should be able to see the output. 

```{r}
#| echo: false
#| eval: true

print("Set this way, the code shouldnot display, but the output should be visible on render")

```

#### Codeblock changes next time through

**Task**
* Change the first codeblock to echo true and eval true. Rewrite the print statement based on what you expect to happen.
* Change the second codeblock to echo false and eval false. Rewrite the print statement based on what you expect to happen.

**Task**: What did you expect to happen? Describe here what happened when you made these changes. 

