# Geographic data analysis

We're going to continue our work on geographic analysis by looking at how these methods can help us answer a slightly harder question. 

Our motivating question for this chapter is this: how many homicides occur near schools in Washington, D.C.?

## Loading libraries

Load libraries as usual. 

```{r}

#install.packages('sf')
#install.packages('leaflet')

library(tidyverse)
library(janitor)
library(sf)

```

## Data

We're loading two dataframes to answer our question.

* Washington Homicides | spatial points | location and details about each Washington homicide between 2007 and 2016. Details on the data: https://github.com/washingtonpost/data-homicides

* Washington Schools | spatial points | locations of each Washington, D.C. school.  https://nces.ed.gov/programs/edge/geographic/schoollocations

* Washington D.C. shapefile | shapefile | outline of Washington, D.C.

```{r}


###
# Load dataframe of washington homicides
###

washington_homicides <- read_rds("data/washington_homicides.rds")

###
# Load dataframe of washington schools
###

washington_schools <- read_rds("data/washington_schools.rds")

###
# Load outline of DC
###

washington_shapefile <- read_rds("data/washington_shapefile.rds")



```

Here are the 239 schools in Washington. 

```{r}
ggplot() +
  geom_sf(data=washington_shapefile, fill="white") + 
  geom_sf(data=washington_schools, color="purple")


```

Here are the 1345 homicides in Washington between 2007 and 2017. 

```{r}
ggplot() +
  geom_sf(data=washington_shapefile, fill="white") + 
  geom_sf(data=washington_homicides, color="red")

```

And here are both on a map.

```{r}
ggplot() +
  geom_sf(data=washington_shapefile, fill="white") + 
  geom_sf(data=washington_homicides, color="red") +
  geom_sf(data=washington_schools, color="purple") 

```

Let's think back to our question: how many homicides occurred near schools?  

In the prior lab, we used something called a "spatial join" to tie a school -- a geographic point -- to a Census tract -- a geographic polygon.  

We could try something similar here. The code below says: create a new object called `schools_homicides` by attempting to connect each point in our schools table with each point in our homicides table.  If there's an overlap, return that match to our new table. 

```{r}

schools_homicides <- washington_schools %>%
  st_join(washington_homicides, left="FALSE")

schools_homicides

```
Notice that it returned zero rows.  That's because for there to be a match, a school point had to be essentially on top of a shooting point.  The points are very small, and it's unlikely there will be an exact match in our data. 

We can fix that by making the size of the point bigger, by converting it to a polygon.  How do we define "near" for the purpose of our question? Let's use "within 200 meters" of the centerpoint of the school as our definition. 

We're going to use a new function `st_buffer` to create a "buffer" around our latitude longitude point.  We're going to change our geometry from a small point to a circle with a radius of 200 meters centered on that point. 

```{r}

washington_schools_buffered <- washington_schools %>%
  mutate(geometry = st_buffer(geometry, dist=100))

```

Let's visualize it. It's hard to see the difference from this map, but those purple circles represent a 200 meter radius circle around our school point.

```{r}

ggplot() +
  geom_sf(data=washington_shapefile, fill="white") + 
  geom_sf(data=washington_schools_buffered, color="purple") 

```

It's easier to see with a leaflet map that shows both the points and the buffered circle.  Zoom in. 
```{r}


#library(leaflet)
leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data=washington_schools_buffered, weight=1, fill="purple") %>%
  addCircles(data=washington_schools,
             label=washington_schools$school_name)
  


```

We can also see the difference in the geometry column between our buffered dataframe and our point dataframe. 

Our washington_schools dataframe is a collection of points. 

```{r}

washington_schools$geometry

```
Our washington_schools_buffered dataframe is a collection of polygons
```{r}

washington_schools_buffered$geometry

```

Now that we have a buffered schools file, we can redo our spatial join.

```{r}
schools_homicides <- washington_schools_buffered %>%
  st_join(washington_homicides, left="FALSE")

schools_homicides

```

We get 87 rows.  This does NOT mean:

* We had 87 homicides near schools.  Some individual homicides occurred within 100 meters of more than one school.
* There were homicides near 87 schools.  Some schools had multiple homicides. 

To answer our original question, "how many homicides occurred near (within 100 meters) of a DC school?", we need to distinct our list of schools and count.  

```{r}
schools_homicides %>%
  distinct(school_name) %>%
  count()
```
There's our answer.  Out of 239 DC schools, 65 -- more than 25 percent -- had a homicide occur within 100 meters.


**Question #1: **
Answer this question in English: Write in Elms

Develop two questions that you could answer with a spatial buffer as described in this chapter, outside of the domain of education and crime.  Briefly describe the two geographic dataframes you would need to answer each question.   




