# Intro to Data Analysis and Visualization in R 

## Step 1: Load libraries

Load a package of code to make doing data journalism easier. 

```{r}

library(tidyverse)
```

## Step 2: Load and examine data

Load the Washington Post police shootings database.

```{r}

police_shootings <- read_csv("https://raw.githubusercontent.com/washingtonpost/data-police-shootings/master/fatal-police-shootings-data.csv")

```

## Step 3: Analyze data

Our question: Which racial group suffered the most shootings by police between 2015 and 2022, according to the Washington Post data? 

Let's make a "pivot table."

```{r}
shootings_by_race <- police_shootings %>%
  group_by(race) %>%
  summarize(
    count_shootings=n()
  ) %>%
  arrange(desc(count_shootings))

shootings_by_race
```

## Step 4: Visualize data

Let's make a simple (and misleading in a big picture truth sense) visualization.

```{r}
shootings_by_race %>%
  ggplot(aes(x=reorder(race,-count_shootings), y=count_shootings)) +
  geom_bar(stat="identity", fill="light blue") +
  labs(y= "Police shootings", x = "Racial group of victim", title="Most people shot by police are white, analysis shows", subtitle="source: Washington Post police shootings data")

```


## Step 5: Make a web page

